---
title: "Constructing g-computation estimators: two case studies in selection bias"
output:
  html_document:
    df_print: paged
---

R code illustrating the application of the proposed g-computation estimators to 
a single data set in the context of the case studies described in the paper.

Haidong Lu

```{r package}
library(data.table)
library(geex)
library(dplyr)
```

#Case 1: 
##Standard g-computation
With treatment-induced selection bias, standard g-computation is expected to be 
biased, as detailed in the paper
```{r}
d <- fread("example1.csv")
d[, AW := A * W]

da1 <- copy(d)
da1[, A := 1]
da1[, AW := A * W]
da0 <- copy(d)
da0[, A := 0]
da0[, AW := A * W]

X <- as.matrix(d[, .(I, A, W, AW)])
X1 <- as.matrix(da1[, .(I, A, W, AW)])
X0 <- as.matrix(da0[, .(I, A, W, AW)])
```


```{r}
inv_logit <- function(x) {
  1 / (1 + exp(-x))
}
```

```{r}
data_df <- as.data.frame(cbind(
  Y = d$Y,
  S = d$S,
  X,
  X1_1 = X1[,1], X1_2 = X1[,2], X1_3 = X1[,3], X1_4 = X1[,4],
  X0_1 = X0[,1], X0_2 = X0[,2], X0_3 = X0[,3], X0_4 = X0[,4]
))
```


```{r}
psi_sg <- function(data) {
  function(theta) {
    # Subset parameters for easier tracking in code later
    rd <- theta[1]
    r1 <- theta[2]
    r0 <- theta[3]
    beta <- theta[4:7]

    y <- data$Y
    s <- data$S
    X <- as.matrix(data[, c("I", "A", "W", "AW")])
    X1 <- as.matrix(data[, c("X1_1", "X1_2", "X1_3", "X1_4")])
    X0 <- as.matrix(data[, c("X0_1", "X0_2", "X0_3", "X0_4")])
    
    # Outcome nuisance model
    Xb <- as.vector(X %*% beta)
    ee_out <- s * (y - inv_logit(Xb)) * X
    
    y1hat <- inv_logit(as.vector(X1 %*% beta))
    y0hat <- inv_logit(as.vector(X0 %*% beta))

    ee_r1 <- y1hat - r1
    ee_r0 <- y0hat - r0
    ee_rd <- rep(r1 - r0 - rd, length(y))

    return(c(ee_rd, ee_r1, ee_r0, as.vector(t(ee_out))))
  }
}
```


```{r}
init_vals <- c(0.0, 0.5, 0.5, 0, 0, 0, 0)

result <- m_estimate(
  estFUN = psi_sg,
  data = data_df,
  root_control = setup_root_control(start = init_vals)
)

estimates <- coef(result)

# calculate covariance
vcov_mat <- vcov(result)
se <- sqrt(diag(vcov_mat))
lower <- estimates - 1.96 * se
upper <- estimates + 1.96 * se


cat("RD estimate:", estimates[1], "\n")
cat("95% CI: [", lower[1], ", ", upper[1], "]\n")

```

##Modified g-computation

```{r}
psi_sg <- function(data) {
  function(theta) {
    # Subset parameters for easier tracking in code later
    rd <- theta[1]
    r1 <- theta[2]
    r0 <- theta[3]
    beta <- theta[4:7]

    y <- data$Y
    s <- data$S
    a <- data$A
    X <- as.matrix(data[, c("I", "A", "W", "AW")])
    X1 <- as.matrix(data[, c("X1_1", "X1_2", "X1_3", "X1_4")])
    X0 <- as.matrix(data[, c("X0_1", "X0_2", "X0_3", "X0_4")])
    
    # Outcome nuisance model
    Xb <- as.vector(X %*% beta)
    ee_out <- s * (y - inv_logit(Xb)) * X
    
    y1hat <- inv_logit(as.vector(X1 %*% beta))
    y0hat <- inv_logit(as.vector(X0 %*% beta))

    ee_r1 <- a * (y1hat - r1)
    ee_r0 <- (1 - a) * (y0hat - r0)
    ee_rd <- rep(r1 - r0 - rd, length(y))

    return(c(ee_rd, ee_r1, ee_r0, as.vector(t(ee_out))))
  }
}
```


```{r}
init_vals <- c(0.0, 0.5, 0.5, 0, 0, 0, 0)

result <- m_estimate(
  estFUN = psi_sg,
  data = data_df,
  root_control = setup_root_control(start = init_vals)
)

estimates <- coef(result)

# calculate covariance
vcov_mat <- vcov(result)
se <- sqrt(diag(vcov_mat))
lower <- estimates - 1.96 * se
upper <- estimates + 1.96 * se


cat("RD estimate:", estimates[1], "\n")
cat("95% CI: [", lower[1], ", ", upper[1], "]\n")

```


#Case 2: Confounding and Selection Bias
##Standard g-computation: X-only
```{r}
d <- fread("example2.csv")

da1 <- copy(d)
da1[, A := 1]
da0 <- copy(d)
da0[, A := 0]

X  <- as.matrix(d[, .(I, A, X)])
X1 <- as.matrix(da1[, .(I, A, X)])
X0 <- as.matrix(da0[, .(I, A, X)])
```

```{r}
data_df <- as.data.frame(cbind(
  Y = d$Y,
  S = d$S,
  X,
  X1_1 = X1[,1], X1_2 = X1[,2], X1_3 = X1[,3],
  X0_1 = X0[,1], X0_2 = X0[,2], X0_3 = X0[,3]
))
```

```{r}
psi_sg <- function(data) {
  function(theta) {
    rd <- theta[1]
    r1 <- theta[2]
    r0 <- theta[3]
    beta <- theta[4:6]

    y <- data$Y; s <- data$S
    X <- as.matrix(data[, c("I", "A", "X")])
    X1 <- as.matrix(data[, c("X1_1", "X1_2", "X1_3")])
    X0 <- as.matrix(data[, c("X0_1", "X0_2", "X0_3")])
    
    Xb <- as.vector(X %*% beta)
    ee_out <- s * (y - inv_logit(Xb)) * X
    
    y1hat <- inv_logit(as.vector(X1 %*% beta))
    y0hat <- inv_logit(as.vector(X0 %*% beta))

    ee_r1 <- y1hat - r1
    ee_r0 <- y0hat - r0
    ee_rd <- rep(r1 - r0 - rd, length(y))

    return(c(ee_rd, ee_r1, ee_r0, as.vector(t(ee_out))))
  }
}
```

```{r}
init_vals <- c(0., 0.5, 0.5, 0., 0., 0.)
result <- m_estimate(
  estFUN = psi_sg,
  data = data_df,
  root_control = setup_root_control(start = init_vals)
)

estimates <- coef(result)
se <- sqrt(diag(vcov(result)))
lower <- estimates - 1.96 * se
upper <- estimates + 1.96 * se

cat("Standard g-computation: X-only\n")
cat("RD:", estimates[1], "\n")
cat("95% CI: [", lower[1], ", ", upper[1], "]\n")
```

##Standard g-computation: Z-only
```{r}
d <- fread("example2.csv")

da1 <- copy(d)
da1[, A := 1]
da0 <- copy(d)
da0[, A := 0]

X  <- as.matrix(d[, .(I, A, Z)])
X1 <- as.matrix(da1[, .(I, A, Z)])
X0 <- as.matrix(da0[, .(I, A, Z)])
```

```{r}
data_df <- as.data.frame(cbind(
  Y = d$Y,
  S = d$S,
  X,
  X1_1 = X1[,1], X1_2 = X1[,2], X1_3 = X1[,3],
  X0_1 = X0[,1], X0_2 = X0[,2], X0_3 = X0[,3]
))
```

```{r}
psi_sg <- function(data) {
  function(theta) {
    rd <- theta[1]
    r1 <- theta[2]
    r0 <- theta[3]
    beta <- theta[4:6]

    y <- data$Y; s <- data$S
    X <- as.matrix(data[, c("I", "A", "Z")])
    X1 <- as.matrix(data[, c("X1_1", "X1_2", "X1_3")])
    X0 <- as.matrix(data[, c("X0_1", "X0_2", "X0_3")])
    
    Xb <- as.vector(X %*% beta)
    ee_out <- s * (y - inv_logit(Xb)) * X
    
    y1hat <- inv_logit(as.vector(X1 %*% beta))
    y0hat <- inv_logit(as.vector(X0 %*% beta))

    ee_r1 <- y1hat - r1
    ee_r0 <- y0hat - r0
    ee_rd <- rep(r1 - r0 - rd, length(y))

    return(c(ee_rd, ee_r1, ee_r0, as.vector(t(ee_out))))
  }
}
```

```{r}
init_vals <- c(0., 0.5, 0.5, 0., 0., 0.)
result <- m_estimate(
  estFUN = psi_sg,
  data = data_df,
  root_control = setup_root_control(start = init_vals)
)

estimates <- coef(result)
se <- sqrt(diag(vcov(result)))
lower <- estimates - 1.96 * se
upper <- estimates + 1.96 * se

cat("Standard g-computation: Z-only\n")
cat("RD:", estimates[1], "\n")
cat("95% CI: [", lower[1], ", ", upper[1], "]\n")
```

##Standard g-computation: X & Z
```{r}
d <- fread("example2.csv")

da1 <- copy(d)
da1[, A := 1]
da0 <- copy(d)
da0[, A := 0]

X  <- as.matrix(d[, .(I, A, X, Z)])
X1 <- as.matrix(da1[, .(I, A, X, Z)])
X0 <- as.matrix(da0[, .(I, A, X, Z)])
```

```{r}
data_df <- as.data.frame(cbind(
  Y = d$Y,
  S = d$S,
  X,
  X1_1 = X1[,1], X1_2 = X1[,2], X1_3 = X1[,3], X1_4 = X1[,4],
  X0_1 = X0[,1], X0_2 = X0[,2], X0_3 = X0[,3], X0_4 = X0[,4]
))
```

```{r}
psi_sg <- function(data) {
  function(theta) {
    rd <- theta[1]
    r1 <- theta[2]
    r0 <- theta[3]
    beta <- theta[4:7]

    y <- data$Y; s <- data$S
    X <- as.matrix(data[, c("I", "A", "X", "Z")])
    X1 <- as.matrix(data[, c("X1_1", "X1_2", "X1_3", "X1_4")])
    X0 <- as.matrix(data[, c("X0_1", "X0_2", "X0_3", "X0_4")])
    
    Xb <- as.vector(X %*% beta)
    ee_out <- s * (y - inv_logit(Xb)) * X
    
    y1hat <- inv_logit(as.vector(X1 %*% beta))
    y0hat <- inv_logit(as.vector(X0 %*% beta))

    ee_r1 <- y1hat - r1
    ee_r0 <- y0hat - r0
    ee_rd <- rep(r1 - r0 - rd, length(y))

    return(c(ee_rd, ee_r1, ee_r0, as.vector(t(ee_out))))
  }
}
```

```{r}
init_vals <- c(0., 0.5, 0.5, 0., 0., 0., 0.)
result <- m_estimate(
  estFUN = psi_sg,
  data = data_df,
  root_control = setup_root_control(start = init_vals)
)

estimates <- coef(result)
se <- sqrt(diag(vcov(result)))
lower <- estimates - 1.96 * se
upper <- estimates + 1.96 * se

cat("Standard g-computation: X + Z\n")
cat("RD:", estimates[1], "\n")
cat("95% CI: [", lower[1], ", ", upper[1], "]\n")
```

##Nested g-computation

```{r}
d <- fread("example2.csv")

da1 <- copy(d)
da1[, A := 1]
da0 <- copy(d)
da0[, A := 0]

X  <- as.matrix(d[, .(I, A, X, Z)])
X1 <- as.matrix(da1[, .(I, A, X, Z)])
X0 <- as.matrix(da0[, .(I, A, X, Z)])

W  <- as.matrix(d[, .(I, A, Z)])
W1 <- as.matrix(da1[, .(I, A, Z)])
W0 <- as.matrix(da0[, .(I, A, Z)])
```


```{r}
data_df <- as.data.frame(cbind(
  Y = d$Y, S = d$S,
  X, W,
  X1_1 = X1[,1], X1_2 = X1[,2], X1_3 = X1[,3], X1_4 = X1[,4],
  X0_1 = X0[,1], X0_2 = X0[,2], X0_3 = X0[,3], X0_4 = X0[,4],
  W1_1 = W1[,1], W1_2 = W1[,2], W1_3 = W1[,3],
  W0_1 = W0[,1], W0_2 = W0[,2], W0_3 = W0[,3]
))
```


```{r}
psi_gcomp_nested <- function(data) {
  function(theta) {
    pX <- 4
    pW <- 3

    rd <- theta[1]
    r1 <- theta[2]
    r0 <- theta[3]
    beta     <- theta[4:(3+pX)]
    gamma_1  <- theta[(4+pX):(3+pX+pW)]
    gamma_0  <- theta[(4+pX+pW):(3+pX+2*pW)]

    y <- data$Y
    s <- data$S

    X  <- as.matrix(data[, c("I", "A", "X", "Z")])
    X1 <- as.matrix(data[, c("X1_1", "X1_2", "X1_3", "X1_4")])
    X0 <- as.matrix(data[, c("X0_1", "X0_2", "X0_3", "X0_4")])

    W  <- as.matrix(data[, c("I", "A", "Z")])
    W1 <- as.matrix(data[, c("W1_1", "W1_2", "W1_3")])
    W0 <- as.matrix(data[, c("W0_1", "W0_2", "W0_3")])

    # Inner model: E[Y | A, X, Z], fit only when S == 1
    Xb <- as.vector(X %*% beta)
    ee_inner <- s * (y - inv_logit(Xb)) * X

    y1hat <- inv_logit(as.vector(X1 %*% beta))
    y0hat <- inv_logit(as.vector(X0 %*% beta))

    # Outer model: E[Y1 | Z], E[Y0 | Z]
    ee_outer1 <- (y1hat - inv_logit(as.vector(W %*% gamma_1))) * W
    ee_outer0 <- (y0hat - inv_logit(as.vector(W %*% gamma_0))) * W

    y1_outer <- inv_logit(as.vector(W1 %*% gamma_1))
    y0_outer <- inv_logit(as.vector(W0 %*% gamma_0))

    # Risk functions
    ee_r1 <- y1_outer - r1
    ee_r0 <- y0_outer - r0
    ee_rd <- rep(r1 - r0 - rd, length(y))

    return(c(ee_rd, ee_r1, ee_r0, as.vector(t(ee_inner)), as.vector(t(ee_outer1)), as.vector(t(ee_outer0))))
  }
}
```


```{r}
init_vals <- c(0., 0.5, 0.5, 0., 0., 0., 0., 0., 0., 0., 0., 0., 0.)

result <- m_estimate(
  estFUN = psi_gcomp_nested,
  data = data_df,
  root_control = setup_root_control(start = init_vals)
)

estimates <- coef(result)
se <- sqrt(diag(vcov(result)))
lower <- estimates - 1.96 * se
upper <- estimates + 1.96 * se

cat("Nested g-computation\n")
cat("RD:", estimates[1], "\n")
cat("95% CI: [", lower[1], ", ", upper[1], "]\n")

```



